Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-10-19 manu-draft.qmd
default results and text added for RN comments

@@

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

By optimization period, median NSE values across all model outputs increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical across optimization periods and median NSE values across model outputs for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all model outputs were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period. About half of the model outputs regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the same combination was observed for the 30 day optimization period with the exception that $\sigma$ was high for $R$. Notably, many of the top models for each optimization period had at least one high value for $\sigma$ for $a$ or $R$, suggesting that models with sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.

The actual model results for the best and worst performing combinations of prior distributions for the three optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all model outputs in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the model outputs were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period. The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplots (b), (d), and (f). In such cases, the $DO$ simulation (fourth row) can still be quite good as a result of compensating errors in $a$ and $R$. The EBASE-derived rates of photosynthesis and respiration are both much higher than in the synthetic data for subplots (b) and (d), errors that compensate well enough to produce fairly accurate recoveries of the $DO$ time series. Further, the worst model for the 30 day optimization period in subplot (f) had the same priors as the 7 day optimization period, but was able to better reproduce the synthetic results, suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Compared to the previous analyses, the prior distribution for the $b$ parameter was not fixed given that its estimate should vary as a function of the data using real observations. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

>>>>

The synthetic time series was used as input to EBASE to predict metabolic parameters with the default prior distributions above. Two optimization periods of 7 and 30 days were evaluated. This analysis provided a demonstration that the default priors are not entirely unreasonable based on the ability to reproduce the known parameters in the synthetic time series [@fig-defplo]. Overall, EBASE produced reasonable results for $a$ and $R$, with higher accuracy using the 30 day optimization period (Figure [-@fig-defplo]b) compared to the 7 day optimization period (Figure [-@fig-defplo]a).  However, the model was unable to return the $b$ parameter with acceptable accuracy, suggesting the data were insufficient to constrain this parameter.  Further, many of the 95% credible intervals from the posterior distributions of each parameter did not overlap the known values from the synthetic time series.  Estimates for each optimization period may be inaccurate despite overall similarities comparing all estimates to the known values for the whole year.

@@

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period. Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series. For the models with 7 and 30 day optimization periods, the final period in the annual time series was excluded from the results due to incomplete number of observations compared to earlier periods (i.e., the total number of days in the time series was not evenly divided by 7 or 30, resulting in a remainder for the last period). Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

By optimization period, median NSE values across all model outputs increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical for the 1 and 7 day optimization period, whereas the top three models for the 30 day optimization appeared less dependent on the choice of priors. Median NSE values across model outputs for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all model outputs were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period. Less than half of the model outputs regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the combination of priors did not have a notable effect on the output for the 30 day optimization period. Notably, many of the top models for the shorter optimization periods had at least one high value for $\sigma$ for $a$ or $R$, suggesting that models with sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.

The actual model results for the best and worst performing combinations of prior distributions for the three optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all model outputs in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the model outputs were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period. The worst performing models for the 1 and 7 day optimization periods (@fig-optex b, d) demonstrate the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplots (b) and (d). In such cases, the $DO$ simulation (fourth row) can still be quite good as a result of compensating errors in $a$ and $R$. The EBASE-derived rates of photosynthesis and respiration are both much higher than in the synthetic data for subplots (b) and (d), errors that compensate well enough to produce fairly accurate recoveries of the $DO$ time series. Further, resutls for the worst model for the 30 day optimization period in subplot (f) are visually indistinguishable from those of the best model in subplot (e), suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

![Comparison of EBASE results with default prior distributions to a synthetic time series created with known model parameters ($a$, $R$, and $b$).  Results are compared for EBASE parameters estimated with a (a) 7 day and (b) 30 day optimization period. 95% credible intervals from the posterior distributions of the parameter estimates from EBASE are also shown. The results are averaged at the time step of the model optimization period.](figs/defplo.png){#fig-defplo}
